---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300,
  fig.width = 5,
  fig.height = 2.5,
  fig.path = "man/figures/README-"
)
```

# tidyplots <a href="https://jbengler.github.io/tidyplots/"><img src="man/figures/logo.svg" align="right" height="139" alt="tidyplots website" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/jbengler/tidyplots/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jbengler/tidyplots/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of `tidyplots` is to streamline the creation of publication-ready plots for scientific papers, making it incredibly easy to gradually add and refine plot elements. It allows precise control over composition, style, and absolute sizes, while its utilization of the pipe `%>%` simplifies the construction of advanced plotting pipelines.

## Disclaimer

This package is still in early development. Expect user-facing and breaking changes. I would not recommend to use this package in production yet.

Feel free to let me know about bugs and feature requests [here](https://github.com/jbengler/tidyplots/issues).

## Installation

``` r
# install.packages("devtools")
devtools::install_github("jbengler/tidyplots")
```

## Usage

Here are some examples of what you can do with just a few key strokes. The full documentation can be found [here](https://jbengler.github.io/tidyplots/).

```{r}
library(tidyplots)

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_data_points_beeswarm()

```

```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute()

```

```{r}
energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_source) %>% 
  add_donut() %>% 
  split_plot(by = year)

```

```{r}
energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_absolute()

```

```{r}
energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_relative()

```

```{r}
study %>% 
  tidyplot(x = group, y = score, color = dose) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_mean_dash() %>% 
  add_mean_value()

```

```{r}
time_course %>%
  tidyplot(x = day, y = score, color = treatment, dodge_width = 0) %>%
  add_mean_line() %>%
  add_mean_dot() %>%
  add_error_ribbon()

```

```{r, fig.height=3}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_boxplot() %>% 
  add_stats_pvalue(ref.group = 1)

```

```{r, fig.height=3.2}
gene_expression %>% 
  dplyr::filter(external_gene_name %in% c("Apol6", "Col5a3", "Vgf", "Bsn")) %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_data_points_beeswarm() %>% 
  add_stats_asterisks(include_info = FALSE) %>% 
  remove_x_axis_title() %>% 
  split_plot(by = external_gene_name)

```

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_data_points_beeswarm() %>% 
  view_plot(title = "Default color scheme: metro") %>% 
  adjust_colors(colors_discrete_candy) %>% 
  view_plot(title = "Alternative color scheme: candy") %>% 
  adjust_colors(colors_discrete_seaside) %>% 
  view_plot(title = "Alternative color scheme: seaside")

```

## Acknowledgements

tidyplots relies on a number of fantastic packages that do all the heavy lifting behind the scenes. These include 
    cli,
    dplyr,
    forcats,
    ggplot2,
    ggpubr,
    ggrastr,
    ggrepel,
    glue,
    Hmisc,
    patchwork,
    purrr,
    ragg,
    rlang,
    scales,
    stringr,
    svglite and
    tidyr.

