---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300,
  fig.width = 5,
  fig.height = 2.5,
  fig.path = "man/figures/README-"
)
```

# tidyplots <a href="https://jbengler.github.io/tidyplots/"><img src="man/figures/logo.svg" align="right" height="139" alt="tidyplots website" /></a>

<!-- badges: start -->
<!-- badges: end -->

tidyplots streamlines the creation of publication-ready plots for scientific papers, making it incredibly easy to incorporate and refine plot elements. It allows precise control over composition, style, and absolute sizes, while its utilization of the pipe `%>%` simplifies the construction of powerful plotting pipelines.

## Installation

``` r
# install.packages("pak")
pak::pak("jbengler/tidyplots")
```

## Usage

Here are some examples of what you can do with 3--5 lines of code.

```{r}
library(tidyplots)

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_jitter()

```

```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute()

```

```{r}
energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_source) %>% 
  add_donut() %>% 
  split_plot(by = year)

```

```{r}
energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_absolute()

```

```{r}
study %>% 
  tidyplot(x = group, y = score, color = dose) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_mean_dash() %>% 
  add_mean_value()

```

```{r}
time_course %>%
  tidyplot(x = day, y = score, color = treatment, dodge_width = 0) %>%
  add_mean_line() %>%
  add_mean_dot() %>%
  add_error_ribbon()

```

```{r, fig.height=3}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_box() %>% 
  add_stats_pvalue(ref.group = 1)

```

## Documentation

https://jbengler.github.io/tidyplots/
