---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300,
  fig.width = 5,
  fig.height = 2.5,
  fig.path = "man/figures/README-"
)
```

# tidyplots

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

## Overview

tidyplots is an opinionated plotting package that makes to it ridiculously simple to create ready-to-use plots for scientific papers. It works by gradually adding and adjusting plot components and provides full control over composition, style, and absolute sizes. tidyplots is powered by the pipe `%>%` and provides a clean and minimalist interface for commonly used scientific plotting routines.

## Installation

You can install the development version of `tidyplots` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jbengler/tidyplots")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(tidyplots)

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_jitter()

```

## Learn more

https://jbengler.github.io/tidyplots/
