---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300,
  fig.width = 5,
  fig.height = 2.5,
  fig.path = "man/figures/README-"
)
```

# tidyplots <a href="https://jbengler.github.io/tidyplots/"><img src="man/figures/logo.svg" align="right" height="139" alt="tidyplots website" /></a>

<!-- badges: start -->
[![R-CMD-check](https://github.com/jbengler/tidyplots/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jbengler/tidyplots/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `tidyplots` is to streamline the creation of publication-ready plots for scientific papers. It allows to gradually add and refine plot elements, and provides precise control over composition, style, and absolute sizes. By consistent use of the pipe `%>%` it simplifies the construction of advanced plotting pipelines.

## Disclaimer

This package is still in early development. Expect user-facing and breaking changes. I would not recommend to use this package in production yet.

Feel free to report bugs and suggest features [here](https://github.com/jbengler/tidyplots/issues).

## Installation

``` r
# install.packages("devtools")
devtools::install_github("jbengler/tidyplots")
```

## Usage

Here are some examples. The full documentation can be found [here](https://jbengler.github.io/tidyplots/).

```{r}
library(tidyplots)

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_sem_bar() %>% 
  add_data_points_beeswarm()

```

```{r}
energy %>% 
  tidyplot(x = year, y = power, color = energy_source) %>% 
  add_barstack_absolute()

```

```{r}
energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_source) %>% 
  add_donut() %>% 
  split_plot(by = year)

```

```{r}
energy_week %>% 
  tidyplot(x = date, y = power, color = energy_source) %>% 
  add_areastack_absolute()

```

```{r}
energy_week %>% 
  tidyplot(x = date, y = power, color = energy_source) %>% 
  add_areastack_relative()

```

```{r}
study %>% 
  tidyplot(x = group, y = score, color = dose) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_mean_dash() %>% 
  add_mean_value()

```

```{r}
time_course %>%
  tidyplot(x = day, y = score, color = treatment, dodge_width = 0) %>%
  add_mean_line() %>%
  add_mean_dot() %>%
  add_sem_ribbon()

```

```{r, fig.height=3}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_boxplot() %>% 
  add_stats_pvalue(ref.group = 1)

```

```{r, fig.height=3.2}
gene_expression %>% 
  dplyr::filter(external_gene_name %in% c("Apol6", "Col5a3", "Vgf", "Bsn")) %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_sem_bar() %>% 
  add_data_points_beeswarm() %>% 
  add_stats_asterisks(include_info = FALSE) %>% 
  remove_x_axis_title() %>% 
  split_plot(by = external_gene_name)

```

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_sem_bar() %>% 
  add_data_points_beeswarm() %>% 
  view_plot(title = "Default color scheme: metro") %>% 
  adjust_colors(colors_discrete_seaside) %>% 
  view_plot(title = "Alternative color scheme: seaside")

```

## Acknowledgements

tidyplots relies on a number of fantastic packages that do all the heavy lifting behind the scenes. These include 
    cli,
    dplyr,
    forcats,
    ggbeeswarm,
    ggplot2,
    ggpubr,
    ggrastr,
    ggrepel,
    glue,
    Hmisc,
    htmltools,
    patchwork,
    purrr,
    rlang,
    scales,
    stringr,
    tidyr, and
    tidyselect.

