---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyplots

<!-- badges: start -->
<!-- badges: end -->

The goal of `tidyplots` is make to it ridiculously simple to create scientific plots by gradually adding, changing and adjusting plot components. It follows a similar logic as `ggplot2`, but uses the pipe `%>%` instead of `+`, which allows for more flexible and powerful workflows. Moreover, `tidyplots` is making the most common scientific plotting routines more accessible, by proving a clean and minimalist interface. Finally, you can always `add()` ggplot code, to handle those cases where you need full control.

## Installation

You can install the development version of `tidyplots` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jbengler/tidyplots")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  fig.align = "center",
  fig.dim = c(3, 2),
  dpi = 300,
  out.width = "80%"
)
```

```{r example}
library(tidyplots)

df_demo %>% 
  tidy_plot(x = category, y = value, color = category, fill = category) %>% 
  add_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_jitter()

```

## Learn more

https://jbengler.github.io/tidyplots/
