---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 2.6,
  dpi = 300
)
```

# Design principles

- `tidyplots` aims at delivering ready-to-publish plots, eliminating cumbersome formatting tasks.

- `tidyplots` uses the pipe `%>%` (instead of `+` like `ggplot2`) to build up plots. This way you can seamlessly pipe **into** and **out of** your plot. For example, coming from fram a data wrangeling pipeline you can generate a plot and directly pipe it into `save_plot()`. You can also call `save_plot()` or `view_plot()` in the middle of your pipeline to output intermediate results.

- `tidyplots` tries to reduce the complexity of `ggplot2` by choosing sensible defaults. However, you take more detailed control by via the `...` argument in each function. And if you need to add plain `ggplot2` code, you can do this using `add()` function, which will preserve the `tidyplots` pipeline.

- All plots have absolute dimensions by default, defined in "mm". These dimensions refer to the plotting area, not the entire plot, thus ensuring consistent lengths of `x` any `y` axes. The dimension can be changed with the `width` and `height` parameters, either when creating the plot with `tidyplot()` or later with `adjust_plot_size()`. If you want to restore the `ggplot2` behavior that a plot automatically takes up all available space, set `width` and `height` to `NA`.

- `tidyplots` allows to split plots into subplots by a variable of choice, similar to `ggplot2::facet_wrap()`. However, `split_plot()` provides a more consistent look and allows to distribute subplots across multiple pages.

- `color` and `fill` are always mapped to the same variable. However, you can still do one of the following: (1) reduce the  opacity of fill using the `alpha` or `saturation` arguments, (2) set `color` or `fill` to a constant hex color within an `add_` function, or (3) set `color` or `fill` to `NA` within an `add_` function to to prevent it from being displayed.

# Workflow

The data to be plotted needs to be **tidy** in the sense that (1) each variable must have its own column, (2) each observation must have its own row and (3) each value must have its own cell. To learn more about tidy data, have a look at the excellent book [R for Data Science](https://r4ds.hadley.nz/data-tidy).

To create a plot, you start with the `tidyplot()` function. Then you pipe through `add_` functions, `adjust_` functions, `theme_` functions and `remove_` functions, preferably in that order. `split_plot()` must be last in this sequence and can only be followed by `save_plot()`.

# Standing on the shoulders of giants
    ggplot2 and the rest of the tidyverse by Hadley Wickham et al.
    patchwork by Thomas Lin Pedersen
    ggpubr by Alboukadel Kassambara
    ggrastr by Evan Biederstedt
    ggrepel by Kamil Slowikowski

# Counts & Sums

```{r setup}
library(tidyplots)
```

```{r}
library(ggplot2)
library(patchwork)

study %>% 
  ggplot(aes(x = treatment, y = score, color = treatment, fill = treatment)) +
  stat_summary(fun = mean, geom = "bar", color = NA, width = 0.6, alpha = 0.3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", linewidth = 0.25, width = 0.4) +
  geom_point(position = position_jitterdodge(jitter.width = 0.2)) +
  scale_y_continuous(expand = expansion(c(0, 0.1))) +
  theme_bw() +
      theme(
      panel.border = element_blank(),
      axis.line = element_line(linewidth = 0.25, colour = "black"),
      plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "mm"),
      plot.background = element_rect(fill = NA, colour = NA),
      legend.background = element_rect(fill = NA, colour = NA),
      legend.key = element_rect(fill = NA, colour = NA),
      strip.background = element_rect(fill = NA, colour = NA),
      panel.background = element_rect(fill = NA, colour = NA),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.ticks = element_line(colour = "black", linewidth = 0.25)
    ) +
  plot_layout(widths = unit(50, "mm"), heights = unit(50, "mm"))

```

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_data_points_jitter()
```
