---
title: "tidyplot"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyplot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 2.6,
  dpi = 300
)
```

# Design principles

- `tidyplots` aims at delivering ready-to-publish plots, eliminating cumbersome formatting tasks.

- `tidyplots` uses the pipe `%>%` (instead of `+` like `ggplot2`) to build up plots. This way you can seamlessly pipe **into** and **out of** your plot. For example, coming from fram a data wrangeling pipeline you can generate a plot and directly pipe it into `save_plot()`. You can also call `save_plot()` or `render_plot()` in the middle of your pipeline to output intermediate results.

- `tidyplots` tries to reduce the complexity of `ggplot2` by choosing sensible defaults. However, you take more detailed control by via the `...` argument in each function. And if you need to add plain `ggplot2` code, you can do this using `add()` function, which will preserve the `tidyplots` pipeline.

- All plots have absolute dimensions by default, defined in "mm". These dimensions refer to the plotting area, not the entire plot, thus ensuring consistent lengths of `x` any `y` axes. The dimension can be changed with the `width` and `height` parameters, either when creating the plot with `tidyplot()` or later with `adjust_size()`. If you want to restore the `ggplot2` behavior that a plot automatically takes up all available space, set `width` and `height` to `NA`.

- `tidyplots` allows to split plots into subplots by a variable of choice, similar to `ggplot2::facet_wrap()`. However, `split_plot()` provides a more consistent look and allows to distribute subplots across multiple pages.

- `color` and `fill` are always mapped to the same variable. However, you can still do one of the following: (1) reduce the  opacity of fill using the `alpha` or `saturation` arguments, (2) set `color` or `fill` to a constant hex color within an `add_` function, or (3) set `color` or `fill` to `NA` within an `add_` function to to prevent it from being displayed.

# Workflow

The data to be plotted needs to be **tidy** in the sense that (1) each variable must have its own column, (2) each observation must have its own row and (3) each value must have its own cell. To learn more about tidy data, have a look at the excellent book [R for Data Science](https://r4ds.hadley.nz/data-tidy).

To create a plot, you start with the `tidyplot()` function. Then you pipe through `add_` functions, `adjust_` functions, `theme_` functions and `remove_` functions, preferably in that order. `split_plot()` must be last in this sequence and can only be followed by `save_plot()`.

# Standing on the shoulders of giants
    ggplot2 and the rest of the tidyverse by Hadley Wickham et al.
    patchwork by Thomas Lin Pedersen
    ggpubr by Alboukadel Kassambara
    ggrastr by Evan Biederstedt
    ggrepel by Kamil Slowikowski

# Counts & Sums

```{r setup}
library(tidyplots)
```

```{r}

animals %>%
  tidyplot(family, size, color = family) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_points() %>% 
  adjust_labels(family, sort_by = size)

study %>%
  tidyplot(treatment, score, color = treatment) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_jitter()

study %>%
  tidyplot(treatment, score, color = treatment) %>%
  add_range() %>% 
  add_points()

study %>%
  tidyplot(treatment, score) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_points() %>%
  add_mean_line()

study %>%
  tidyplot(treatment, score) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_points() %>%
  add_line(group = participant, dodge_width = 0)

study %>%
  tidyplot(group, score, color = dose) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_jitter()

p1 <- 
  study %>%
  tidyplot(x = treatment, y = score, color = treatment) %>%
  add_mean_dash() %>%
  add_error() %>%
  add_jitter()
p1

p2 <- 
  study %>%
  tidyplot(x = treatment, y = score, color = treatment) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_jitter()
p2

study %>%
  tidyplot(x = treatment, y = score, color = treatment) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_jitter(style = "confetti")

# testing adjust_colors() adjust_labels() split_plot() interference
# "split" always after "adjust"
# "adjust_color()" always after "adjust_labels()"

# providing less or more colors, not a problem
new_colors <-
  c("A" = "#B0B1B3",
    "B" = "#F18823",
    "C" = "#E23130",
    "D" = "#1D5D83")

# providing less names, not a problem. Providing more names -> Error
new_names <-
  c("A" = "Regime A",
    "B" = "Regime B",
    "C" = "Regime C",
    "D" = "Regime D")

p2 %>% adjust_colors(new_colors)
p2 %>% adjust_labels(treatment, new_order = c("C")) %>% adjust_colors(new_colors)
p2 %>% adjust_colors(new_colors) %>% adjust_labels(treatment, new_order = c("C"))

p2 %>% adjust_colors(new_colors)
p2 %>% adjust_labels(treatment, new_names = new_names) %>% adjust_colors(new_colors)
p2 %>% adjust_colors(new_colors) %>% adjust_labels(treatment, new_names = new_names)

# providing more colors, not a problem. Providing less colors -> Error
new_colors <-
  c("#B0B1B3",
    "#F18823",
    "#E23130",
    "#1D5D83")

p2 %>% adjust_colors(new_colors)
p2 %>% adjust_labels(treatment, new_order = c("C")) %>% adjust_colors(new_colors)
p2 %>% adjust_colors(new_colors) %>% adjust_labels(treatment, new_order = c("C"))

p2 %>% adjust_colors(new_colors)
p2 %>% adjust_labels(treatment, new_names = new_names) %>% adjust_colors(new_colors)
p2 %>% adjust_colors(new_colors) %>% adjust_labels(treatment, new_names = new_names)

# during split plot, single subplots might only have NA for certain categories
# this has been fixed apparently by using drop = FALSE in 

# adjusting axis limits

p1 %>%
  adjust_y_axis(limits = c(-20, 60))

# x--y plots


p3 <- 
  animals %>% 
  tidyplot(x = weight, y = size, color = number_of_legs) %>% 
  add_points(style = "confetti") %>% 
  add_points(style = "confetti", data = filter_rows(size > 300), color = "red") %>% 
  add_text(data = max_rows(weight, n = 3), var = animal, color = "black")
  
p3
p3 %>% add_reference_lines(x = c(4000), y = c(200, 350))

# labels of out of bounds animals still appear
p3 %>%
  adjust_y_axis(limits = c(0, 600)) %>% 
  adjust_x_axis(limits = c(0, 1000))

energy_week %>%
  tidyplot(Date, power_in_mw, color = energy_type) %>%
  add_mean_line() %>%
  adjust_y_axis(limits = c(1000, 5000)) %>% 
  remove_padding()

# adding titles and captions

p2 %>% add_title("My fancy long title")
p2 %>% add_caption("Here goes the fine print")

# adjusting axis titles

p2 %>% adjust_x_axis("My x axis title")
p2 %>% adjust_y_axis("My y axis title")

# adjusting legend

p2 %>% adjust_legend(title = "My legend title")
p2 %>% adjust_legend(title = "My legend title", position = "top")

# plotmath expressions

p2 %>% add_title(title = "$E==m*c^{2}~H[2]*O$")
p2 %>% add_caption(caption = "$E==m*c^{2}~H[2]*O$")
p2 %>% adjust_legend(title = "$E==m*c^{2}~H[2]*O$")
p2 %>% adjust_x_axis(title = "$Domino~E==m*c^{2}$")
p2 %>% adjust_y_axis(title = "$Domino~E==m*c^{2}$")

new_names <-
  c("A" = "$TNF*alpha$",
    "B" = "$IFN*gamma$",
    "C" = "plain text",
    "D" = "$H[2]*O$")

p2 %>% adjust_labels(treatment, new_names = new_names)

# adjust labels

new_names <-
  c("A" = "Regime A",
    "B" = "Regime B",
    "C" = "Regime C",
    "D" = "Regime D")

p2 %>% adjust_labels(treatment, new_names = new_names)
p2 %>% adjust_labels(treatment, new_names = new_names) %>% adjust_x_axis(rotate_labels = TRUE)
p2 %>% adjust_labels(treatment, new_names = new_names, sort_by = score)
p2 %>% adjust_labels(treatment, new_names = new_names, reverse = TRUE)

new_order <- 
  c("D", "C")

p2 %>% adjust_labels(treatment, new_order = new_order)
p2 %>% adjust_labels(treatment, sort_by = score)
p2 %>% adjust_labels(treatment, reverse = TRUE)

p2 %>% adjust_labels(treatment)

# adjust colors

new_colors <-
  c("A" = "#B0B1B3",
    "B" = "#F18823",
    "C" = "#E23130",
    "D" = "#1D5D83")

p2 %>% adjust_colors(new_colors)
p2 %>% adjust_colors(new_colors, as_palette = TRUE)

# new_colors AND new_names

p2 %>%
  adjust_labels(treatment, new_names = new_names) %>% 
  adjust_colors(new_colors, saturation = 0.3)

p2 %>% 
  adjust_colors(new_colors, saturation = 0.3) %>% 
  adjust_labels(treatment, new_names = new_names)

# add_heatmap()

gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap() %>% 
  adjust_size(height = 90)

gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>% 
  adjust_labels(external_gene_name, sort_by = -dplyr::desc(direction)) %>%
  adjust_size(height = 90)

gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row", rasterize = TRUE, rasterize_dpi = 20) %>% 
  adjust_labels(external_gene_name, sort_by = -dplyr::desc(direction)) %>%
  adjust_size(height = 90)

gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "column") %>% 
  adjust_size(height = 90)

# plotmath expressions

p2 %>% adjust_x_axis(title = "$Domino~E==m*c^{2}$")
p2 %>% adjust_y_axis(title = "$Domino~E==m*c^{2}$")

# next chapter

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter(rasterize = TRUE)

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3)

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_mean_dash() %>% 
  add_mean_value()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_jitter()

study %>% 
  tidyplot(x = group, y = score, color = dose) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_jitter()

study %>% 
  tidyplot(x = group, y = score, color = dose) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_mean_dash() %>% 
  add_mean_value()

study %>% 
  tidyplot(x = group, y = score, color = dose, dodge_width = 0.6) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_mean_dash() %>% 
  add_mean_value()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(width = 0.02) %>% 
  add_mean_dot()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_bar(width = 0.02) %>% 
  add_mean_dot() %>% 
  add_mean_value(vjust = -1, padding_top = 0.2)

study %>% 
  tidyplot(x = group, y = score, color = dose) %>% 
  add_mean_bar(width = 0.02) %>% 
  add_mean_dot()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_range() %>% 
  add_mean_dot()

# box and violin

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_range() %>% 
  add_mean_dash()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_box()

study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_box() %>% 
  add_jitter()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_violin() %>% 
  add_jitter()

study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_violin() %>% 
  add_jitter()

# adjust_size

p1
p1 %>% adjust_size(width = 70)
p1 %>% adjust_size(width = 44, height = 25)
p1 %>% adjust_size(width = NA, height = NA)

# line plots

study %>% 
  tidyplot(x = group, y = score, group = participant, color = dose) %>% 
  add_line() %>% 
  add_area(alpha = 0.1)

# summary lines

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_mean_line()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_mean_line() %>% 
  add_mean_dot()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_mean_line() %>% 
  add_mean_dot() %>% 
  add_error()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_mean_line() %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_error_ribbon() %>% 
  add_mean_line() %>% 
  add_jitter()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_range_ribbon() %>% 
  add_mean_line() %>% 
  add_jitter()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_sd_ribbon() %>% 
  add_mean_line() %>% 
  add_jitter()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_ci95_ribbon() %>% 
  add_mean_line() %>% 
  add_jitter()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_curve() %>% 
  add_jitter()

# grouped line

study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_mean_line() %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter()

study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_error_ribbon() %>% 
  add_mean_line() %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter()

study %>% 
  tidyplot(x = dose, y = score, color = group, dodge_width = 0) %>% 
  add_mean_line() %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter()

study %>% 
  tidyplot(x = dose, y = score, color = group, dodge_width = 0) %>% 
  add_error_ribbon() %>% 
  add_mean_line() %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter()

# apparently there can be no lines between doged data points with the same x axis label
# because they need to be in a different aes(group) to be dodged AND they need to in the same aes(group) to be connected by a geom_line()
# solution: bring them to different x axis labels

study %>% 
  tidyplot(x = treatment, y = score, color = group) %>% 
  add_line(group = participant, dodge_width = 0, color = "grey") %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_points()

study %>%
  tidyplot(treatment, score, color = group) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_points() %>%
  add_line(group = participant, dodge_width = 0, color = "grey") %>%
  adjust_labels(treatment, sort_by = dose)

# proportions

animals %>% 
  tidyplot(x = diet, color = family) %>% 
  add_barstack_absolute()

animals %>% 
  tidyplot(x = number_of_legs, color = family) %>% 
  add_barstack_relative()

animals %>% 
  tidyplot(x = family, color = number_of_legs) %>% 
  add_barstack_relative()

animals %>% 
  tidyplot(color = diet) %>% 
  add_donut()

animals %>% 
  tidyplot(color = activity) %>% 
  add_donut()

animals %>% 
  tidyplot(color = number_of_legs) %>% 
  add_donut()

# spendings

spendings %>% 
  tidyplot(color = category) %>% 
  add_donut()

spendings %>% 
  tidyplot(y = amount, color = category) %>% 
  add_barstack_absolute()

spendings %>% 
  tidyplot(y = amount, color = category) %>% 
  add_barstack_relative()

spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar() %>% 
  add_sum_value(accuracy = 1) %>% 
  adjust_x_axis(rotate_labels = TRUE)

spendings %>% 
  tidyplot(x = date, y = amount, color = category) %>% 
  add_sum_bar(preserve = "single")

spendings %>% 
  tidyplot(x = date, y = amount, color = category) %>% 
  add_sum_dot() %>% 
  add_sum_value(accuracy = 1, vjust = -1.2)

spendings %>% 
  dplyr::mutate(cum_amount = cumsum(amount)) %>% 
  tidyplot(x = date, y = cum_amount) %>% 
  add_line() %>% 
  add_points()

spendings %>% 
  dplyr::mutate(cum_amount = cumsum(amount)) %>% 
  tidyplot(x = date, y = cum_amount) %>% 
  add_area(alpha = 0.3) %>% 
  add_points()

spendings %>% 
  dplyr::mutate(cum_amount = cumsum(amount), .by = category) %>% 
  tidyplot(x = date, y = cum_amount, color = category, dodge_width = 0) %>% 
  add_area(alpha = 0.3)

spendings %>% 
  tidyr::complete(date, category, fill = list(title = NA, amount = 0)) %>% 
  dplyr::mutate(cum_amount = cumsum(amount), .by = category) %>% 
  tidyplot(x = date, y = cum_amount, color = category, dodge_width = 0) %>% 
  add_area(alpha = 0.3)

spendings %>% 
  tidyr::complete(date, category, fill = list(title = NA, amount = 0)) %>% 
  dplyr::mutate(cum_amount = cumsum(amount), .by = category) %>% 
  tidyplot(x = date, y = cum_amount, color = category, dodge_width = 0) %>% 
  add_areastack_absolute(alpha = 0.3)

# subset data

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_points()

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_points(data = filter_rows(treatment == "C"))

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_points(data = max_rows(score, n = 1, by = treatment))

# adjust_description

p1 %>% add_title(title = "New title")

# add stats

p2
p2 %>% add_stats_pvalue(ref.group = 1)
p2 %>% add_stats_pvalue(ref.group = 1, p.adjust.method = "bonferroni")
p2 %>% add_stats_asterisks(ref.group = 1, p.adjust.method = "bonferroni")

# grouped

p3 <- 
  study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error() %>% 
  add_jitter()

p3
p3 %>% add_stats_pvalue()
p3 %>% add_stats_pvalue(p.adjust.method = "bonferroni")
p3 %>% add_stats_asterisks(p.adjust.method = "bonferroni")

# animals

animals %>%
  tidyplot(x = family) %>%
  add_count_bar(alpha = 0.3) %>% 
  add_count_value(accuracy = 1) %>% 
  add_count_dash()

animals %>%
  tidyplot(x = family, y = weight) %>%
  add_sum_bar(alpha = 0.3) %>% 
  add_sum_value() %>% 
  add_sum_dash()

animals %>%
  tidyplot(x = family, y = weight) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_mean_value()

animals %>%
  tidyplot(x = family, y = weight) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_mean_value(vjust = 0.4, hjust = -0.2, padding_top = 0.3) %>%
  adjust_flip()

animals %>%
  tidyplot(x = family, y = speed) %>%
  add_mean_bar(alpha = 0.3) %>%
  add_error() %>%
  add_jitter() %>%
  adjust_x_axis(rotate_labels = TRUE)

# split_plot

gene_expression %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter() %>% 
  adjust_x_axis(title = "") %>% 
  split_plot(by = external_gene_name, ncol = 4, nrow = 4)

gene_expression %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter() %>% 
  adjust_x_axis(title = "") %>% 
  split_plot(by = external_gene_name, ncol = 4, nrow = 4, widths = 15, heights = 15)

gene_expression %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter() %>% 
  adjust_x_axis(title = "") %>% 
  split_plot(by = external_gene_name, ncol = 4, nrow = 4, widths = NA, heights = NA)

# stats with split_plot

gene_expression %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter() %>% 
  add_stats_pvalue(include_info = FALSE) %>% 
  adjust_x_axis(title = "") %>% 
  split_plot(by = external_gene_name, ncol = 4, nrow = 4, widths = 25, heights = 25)

gene_expression %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter() %>% 
  add_stats_asterisks(include_info = FALSE) %>% 
  adjust_x_axis(title = "") %>% 
  split_plot(by = external_gene_name, ncol = 4, nrow = 4, widths = 25, heights = 25)

```


```{r eval=FALSE}
# save_plot

gene_expression %>% 
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_error() %>% 
  add_jitter() %>% 
  adjust_x_axis(title = "") %>% 
  split_plot(by = external_gene_name, ncol = 4, nrow = 4) %>%
  save_plot(filename = "test_multipage.pdf")

# save intermediate plots

study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  save_plot(filename = "test1.pdf") %>%
  add_error() %>% 
  save_plot(filename = "test2.pdf") %>% 
  add_jitter() %>% 
  save_plot(filename = "test3.pdf")
```

