---
title: "Color schemes"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 2.5,
  dpi = 300
)

```

::: {.lead}
In this article, we will demonstrate the use of color schemes in tidyplots. We will explore the default color schemes that come with tidyplots and are ready to use for plotting. These include schemes for discrete, continuous and diverging variables. To conclude, we will discuss the creation of custom color schemes from hex values.
:::

# Default color schemes

tidyplots comes with a number of default color schemes. Many of them are adapted from the `viridisLite` and `RColorBrewer` packages. You access them by loading the the tidyplots library and start typing `colors_`. The auto-completion will guide you through a selection of `discrete`, `continuous` and `diverging` schemes.

Let's have a look at the signature scheme of tidyplots, which is called `colors_discrete_metro`. When running the line `colors_discrete_metro` in the console or within a script, a preview of the scheme will be rendered to the Viewer pane in the lower right of the RStudio Desktop interface.

In essence, tidyplots color schemes are just a character vector of hex colors with a special print method that sends a preview to the RStudio viewer pane.

```{r}
library(tidyplots)
colors_discrete_metro
```

```{r results='asis', echo=FALSE}
print(colors_discrete_metro, return_html = TRUE)
```

_Tip: You can copy individual hex colors directly from the preview to use them in your script._

## Discrete

Discrete color schemes are meant for categorical variables. The default schemes in tidyplots consist of 5--6 colors. However, if more categories are present in the plot, tidyplots will automatically fill up the gaps between colors to deliver exactly the number that is required for the plot.

```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute()
```

And here are some variations.

```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_seaside)
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_candy)
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_pastel)
```

## Continuous

Continuous color schemes are meant for continuous variables. The default schemes in tidyplots usually consist of 265 colors.

```{r}
colors_continuous_viridis
```

```{r results='asis', echo=FALSE}
print(colors_continuous_viridis, return_html = TRUE)
```

Here is a use case for a continuous color scheme.

```{r}
climate %>% 
  tidyplot(x = month, y = year, color = max_temperature) %>% 
  add_heatmap()
```

And here are some variations.

```{r}
climate %>% 
  tidyplot(x = month, y = year, color = max_temperature) %>% 
  add_heatmap() %>% 
  adjust_colors(new_colors = colors_continuous_viridis)
climate %>% 
  tidyplot(x = month, y = year, color = max_temperature) %>% 
  add_heatmap() %>% 
  adjust_colors(new_colors = colors_continuous_mako)
climate %>% 
  tidyplot(x = month, y = year, color = max_temperature) %>% 
  add_heatmap() %>% 
  adjust_colors(new_colors = colors_continuous_turbo)
climate %>% 
  tidyplot(x = month, y = year, color = max_temperature) %>% 
  add_heatmap() %>% 
  adjust_colors(new_colors = colors_continuous_rocket)
```

## Diverging

Diverging color schemes are meant for continuous variables that have a central point in the middle. A classical example is the blue--white--red gradient used for gene expression heatmaps.

```{r}
colors_diverging_blue2red
```

```{r results='asis', echo=FALSE}
print(colors_diverging_blue2red, return_html = TRUE)
```

Here is a use case for a diverging color scheme.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  sort_y_axis_labels(direction) %>% 
  adjust_plot_area_size(height = 100)
```

And here are some variations.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  sort_y_axis_labels(direction) %>% 
  adjust_plot_area_size(height = 100) %>% 
  adjust_colors(new_colors = colors_diverging_blue2brown)
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  sort_y_axis_labels(direction) %>% 
  adjust_plot_area_size(height = 100) %>% 
  adjust_colors(new_colors = colors_diverging_BuRd)
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  sort_y_axis_labels(direction) %>% 
  adjust_plot_area_size(height = 100) %>% 
  adjust_colors(new_colors = colors_diverging_BuYlRd)
```

# Custom color schemes

Of course you can also construct custom color schemes using the `new_color_scheme()` function.

```{r}
my_colors <- 
  new_color_scheme(c("#ECA669","#E06681","#8087E2","#E2D269"),
                   name = "my_custom_color_scheme")
my_colors
```

```{r results='asis', echo=FALSE}
print(my_colors, return_html = TRUE)
```

Than you can use your scheme as input to the `adjust_colors()` function.

```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(new_colors = my_colors)
```

Besides creating new schemes, you can also subset and concatenate existing schemes in the exact same way you would do with a regular character string.

```{r}
colors_discrete_metro[2]
```

```{r results='asis', echo=FALSE}
print(colors_discrete_metro[2], return_html = TRUE)
```

```{r}
colors_discrete_metro[2:4]
```

```{r results='asis', echo=FALSE}
print(colors_discrete_metro[2:4], return_html = TRUE)
```

```{r}
c(colors_discrete_metro, colors_discrete_seaside)
```

```{r results='asis', echo=FALSE}
print(c(colors_discrete_metro, colors_discrete_seaside), return_html = TRUE)
```

# What's more?

To dive deeper into code-based plotting, here a couple of resources.

## tidyplots documentation

- [Reference](https://jbengler.github.io/tidyplots/reference/index.html)  
Overview of all tidyplots functions

- [Get started](file:///Users/janbroderengler/GoogleDrive/R/packages/tidyplots/docs/articles/tidyplots.html)  
Getting started guide

- [Visualizing data](https://jbengler.github.io/tidyplots/articles/Visualizing-data.html)  
Article with examples for common data visualizations

- [Advanced plotting](https://jbengler.github.io/tidyplots/articles/Advanced-plotting.html)  
Article about advanced plotting techniques and workflows

- [Color schemes](https://jbengler.github.io/tidyplots/articles/Color-schemes.html)  
Article about the use of color schemes

## Other resources

- [Hands-On Programming with R](https://rstudio-education.github.io/hopr/)   
Free online book by Garrett Grolemund

- [R for Data Science](https://r4ds.hadley.nz)  
Free online book by Hadley Wickham

- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)  
Free online book by Claus O. Wilke
