---
title: "Visualizing data"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 2.5,
  dpi = 300
)

```

::: {.lead}
In this article, we will explore the visualization of different types of datasets using tidyplots. We will cover the plotting of raw data points, amounts, and heatmaps, as well as measures for central tendency, dispersion, and uncertainty. We will conclude by visualizing distributions and proportions, and adding statistical comparisons and annotations.
:::

# Data points

Plotting the raw data points is probably the most bare bone way to visualize a dataset. The corresponding function in tidyplots is called `add_data_points()`.

```{r}
library(tidyplots)

animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points()
```

In the above example some data points appear to overlap other points. To account for this so called _overplotting_, you might want to add a thin white border around the points. This is achieved by setting the argument `white_border = TRUE`.

```{r}
animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points(white_border = TRUE)
```

Another way is to make the points transparent using the `alpha` argument.

```{r}
animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points(alpha = 0.4)
```

Or to change the plotting symbol to an open `shape`.

```{r}
animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points(shape = 1)
```

However, data points can also be used to plot a _discrete variable_ against a _continuous variable_.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points()
```

To avoid overplotting in this scenario, there are two additional options. You can add some random noise to the y position, also known as _jitter_.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points_jitter()
```

Alternatively, you can use an algorithm that keeps the points centered and just moves potentially overlapping points to the sides.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points_beeswarm()
```

# Amounts

For some datasets, it makes sense to `count` or `sum` up data points in order to arrive to conclusions. As one example, let's have a look at the `spendings` dataset.

```{r, results='markup'}
spendings
```

As you can see, this dataset contains family spendings over a time period of 15 days in October. Here, it might be informative to see which spending categories are reoccurring and which are just one time spendings.

```{r}
spendings %>% 
  tidyplot(x = category) %>% 
  add_count_bar()
```

One thing to note here is that the x axis labels are overlapping and are thus unreadable. There are at least two possible solutions for this. One is to swap the x and y axis.

```{r}
spendings %>% 
  tidyplot(y = category) %>% 
  add_count_bar()
```

The other one is to rotate the x axis labels.

```{r}
spendings %>% 
  tidyplot(x = category) %>% 
  add_count_bar() %>% 
  adjust_x_axis(rotate_labels = TRUE)
```

Now we can appreciate that this family had reoccurring spendings for _Food_ but just one spending for _Housing_.

Next, we ask the question _how much_ was spend on each of the categories by plotting the `sum` amount.

```{r}
spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar() %>% 
  adjust_x_axis(rotate_labels = TRUE)
```

Note that we had to introduce the argument `y = amount` in the `tidyplot()` function to make it clear which variable should be summed up.

I also added `color = category` in the `tidyplot()` function to have the variable `category` encoded by different colors.

Since the labels for the variable `category` are now duplicated in the plot, one could argue that it would be justified to remove the duplicated information on the x axis.

```{r}
spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar() %>% 
  adjust_x_axis(rotate_labels = TRUE) %>% 
  remove_x_axis_labels() %>% 
  remove_x_axis_title() %>% 
  remove_x_axis_ticks()
```

Note that besides the x axis labels, I also removed the x axis ticks and x axis title to achieve a cleaner look.

Of course you are free to play around with different graphical representations of the sum values. Here is an example of a lollipop plot constructed from a thin `bar` and a `dot`.

```{r}
spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar(width = 0.03) %>% 
  add_sum_dot() %>% 
  add_sum_value(accuracy = 1) %>% 
  adjust_x_axis(rotate_labels = TRUE) %>% 
  remove_x_axis_labels() %>% 
  remove_x_axis_title() %>% 
  remove_x_axis_ticks()
```

I also added the sum value as text label using the `add_sum_value()` function. 

# Heatmaps

Heatmaps are a great way to plot a _continuous variable_ across two additional _variables_. To exemplify this, we will have a look at the `gene_expression` dataset.

```{r, results='markup'}
gene_expression %>% 
  dplyr::glimpse()
```

We will start by plotting the `expression` values of each `external_gene_name` across the `sample` variable.

```{r}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap()
```

One thing to note here is that the y axis labels are overlapping. So let's increase the height of the plot area from 50 to 100 mm.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap() %>%
  adjust_size(height = 100)
```

The next thing to note is that some of the rows like _Map1a_ and _Kif1a_ show very high values while other rows show much lower values. Let's apply a classical technique to reserve the color range for differences within each row. This is done by calculating _row z scores_ for each row individually. Luckily, tidyplots does this for us when setting the argument `scale = "row"` within the `add_heatmap()` function call.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  adjust_size(height = 100)
```

Now it much easier to appreciate the dynamics of individual genes across the samples on the x axis.

However, the rows appear to be mixed. Some having rather high expression in the "Eip" samples while others have high value in the "Hip" samples. Conveniently, there is a variable called `direction` in the dataset, which classifies genes as being either "up" or "down" regulated. Let's use this variable to sort our y axis.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  adjust_size(height = 100) %>% 
  sort_y_axis_labels(direction)
```

# Central tendency

In cases with multiple data points per experimental group, the `mean` and the `median` are a great way to compute a typical center value for the group, also known as central tendency measure. In tidyplots, these function start with `add_mean_` or `add_median_`.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points_beeswarm() %>% 
  add_mean_dash()
```

The second part of the function name is dedicated to the graphical representation. These include the representation as `bar`, `dash`, `dot`, `value`, `line` or `area`. Of course, these different representations can also be combined. Like in this case `line` and `dot`.

```{r}
time_course %>% 
  tidyplot(x = day, y = score, color = treatment, dodge_width = 0) %>% 
  add_mean_line() %>% 
  add_mean_dot()
```

Or in this case `line` and `area`.

```{r}
time_course %>% 
  tidyplot(x = day, y = score, color = treatment, dodge_width = 0) %>% 
  add_mean_line(linewidth = 1) %>% 
  add_mean_area(alpha = 0.2)
```

In case you ask yourself why I included `dodge_width = 0` in all `tidyplot()` calls used to show a `line` or `area`, you might want to learn more about dodging in the article about [Advanced plotting](https://jbengler.github.io/tidyplots/articles/Advanced-plotting.html#dodging).

Back to central tendency measures, here is one more example using `bar` and `value`.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_mean_bar() %>% 
  add_mean_value()
```

You could also plot the `mean` and the `median` together to explore in which cases they diverge. In the example below the `mean` is shown in orange and the `median` in purple.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points_beeswarm(color = "grey") %>% 
  add_mean_dash(color = "orange") %>% 
  add_median_dash(color = "purple")
```

# Dispersion & uncertainty

To complement the central tendency measure, it is often helpful to provide information about the variability or dispersion of the data points. Such measures include the standard error of the mean `sem`, the standard deviation `sd`, the `range` from the highest to the lowest data point and the 95% confidence interval `ci95`.

A classical representation of dispersion is an `errorbar`.

```{r}
time_course %>% 
  tidyplot(x = day, y = score, color = treatment) %>% 
  add_mean_line() %>% 
  add_mean_dot() %>% 
  add_sem_errorbar(width = 2)
```

Or the use of a semitransparent `ribbon`.

```{r}
time_course %>% 
  tidyplot(x = day, y = score, color = treatment) %>% 
  add_mean_line() %>% 
  add_mean_dot() %>% 
  add_sem_ribbon()
```

Another widely used alternative, especially for not normally distributed data is the use of `violin` or `boxplot`. Starting with the `violin`, the shape of these plots resembles the underlying distribution of the data points.

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_violin()
```

These can be further augmented by adding, for example, the 0.5 quantile and the underlying data points.

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_violin(draw_quantiles = 0.5) %>% 
  add_data_points_beeswarm()
```

The `boxplot` is the more classical approach, in which the quantiles are visualized by a box and whiskers.

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_boxplot()
```

Finally, although it is not strictly a measure of central tendency, you can fit a curve through your data to derive an abstracted representation.

```{r}
time_course %>% 
  tidyplot(x = day, y = score, color = treatment, dodge_width = 0) %>% 
  add_curve_fit()
```

# Distribution

When looking at a single distribution of values, a classical approach for visualization is a `histogram`.

```{r}
energy %>% 
  tidyplot(x = power) %>% 
  add_histogram()
```

To represent the density of values along the x axis another approach is a `density_histogram` or `density_curve`, which also can be combined in one plot.

```{r}
energy %>% 
  tidyplot(x = power) %>% 
  add_density_histogram() %>% 
  add_density_curve()
```

If you want to compare multiple distributions, `violin` or `boxplot` are two potential solutions.

```{r}
distributions %>% 
  tidyplot(x = name, y = value) %>% 
  add_violin() %>% 
  add_data_points_beeswarm()
```

# Proportion

Proportional data provides insights into the proportion or percentage that each individual category contributes to the total. To explore the visualization of proportional data in tidyplots, let's introduce the `energy` dataset.

```{r, results='markup'}
energy %>% 
  dplyr::glimpse()
```

As you might appreciate, this dataset contains the `power` in gigawatt produced per `energy_source` in Germany between `year` 2002 and 2023. Let's start with a `pie` plot.

```{r}
energy %>% 
  tidyplot(color = energy_type) %>% 
  add_pie()
```

The above plot represents the `count` of values across the different `energy_type` categories.

However, we might be more interested, in the `sum` contribution of each `energy_type` to the total `power` production. Therefore, we have to provide the variable `power` as a `y` argument to the `tidyplots()` function.

```{r}
energy %>% 
  tidyplot(y = power, color = energy_type) %>% 
  add_donut()
```

Now we can appreciate the contribution of each energy type. Note that I also changed the `pie` for a `donut` plot, which is basically a pie chart with a white hole in the center.

The main criticism of `pie` and `donut` plots is that the human brain struggles to accurately interpret the proportions represented. For example, it's challenging to determine whether fossil or renewable power has a larger share in the plot.

Another graphical representation is a `barstack` plot.

```{r}
energy %>% 
  tidyplot(y = power, color = energy_type) %>% 
  add_barstack_absolute()
```

Here it might be slightly easier to compare the energy types. However, to really pinpoint this, we probably need to go back to a classical bar plot.

```{r}
energy %>% 
  tidyplot(x = energy_type, y = power) %>% 
  add_sum_bar() %>% 
  add_sum_value() %>% 
  reorder_x_axis_labels("Renewable")
```

Now we can see that renewable power production had a slightly larger overall share than fossil power production in the period between 2002 and 2023.

However, if we want to visualize proportional data across time or another variable, `barstack` plots can still be informative.

```{r}
energy %>% 
  tidyplot(x = year, y = power, color = energy_type) %>% 
  add_barstack_absolute()
```

Moreover, to see the relative instead of the absolute contribution, we can use the `add_barstack_relative()` function.

```{r}
energy %>% 
  tidyplot(x = year, y = power, color = energy_type) %>% 
  add_barstack_relative()
```

A similar plot can be achieved using an `areastack`.

```{r}
energy %>% 
  tidyplot(x = year, y = power, color = energy_type) %>% 
  add_areastack_relative()
```

In both plots, the increasing contribution of renewable energy to the total energy production over time becomes apparent.

This can also be shown using donut plots. However, we need to downsample the dataset to 4 representative years.

```{r}
energy %>% 
  # downsample to 4 representative years
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  # start plotting
  tidyplot(y = power, color = energy_type) %>% 
  add_donut() %>% 
  adjust_colors(new_colors = c("Fossil" = "grey",
                               "Nuclear" = "#F6C54D",
                               "Renewable" = "#4FAE62",
                               "Other" = "#C02D45")) %>%  
  split_plot(by = year)
```

Now, let’s examine a related dataset that presents one week of energy data with higher time resolution.

```{r}
energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_absolute()
```

In this plot, one can appreciate the higher contribution of solar power during day time in comparison to night time.

Also this plot can be shown as a relative areastack.

```{r}
energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_relative()
```

This illustrates nicely how wind energy compensates for the lack of solar power during the night. However, when wind is weak, as on September 10, fossil energy sources need to step in to fill the gap.

# Statistical comparison

To test for differences between experimental groups, tidyplots offers the functions `add_test_asterisks()` and `add_test_pvalue()`. While the first one includes asterisks for symbolizing significance.

```{r}
study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_asterisks()
```

`add_test_pvalue()` provides the computed _p_ value.

```{r}
study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_pvalue()
```

As you might have noted, when using these functions, a caption is automatically included that provides details about the statistical testing performed. The default is a Student's _t_ test without multiple comparison adjustment.

Both can be changed by providing the `method` and `p.adjust.method` arguments.

For example, let's perform a Wilcoxon signed-rank test with Benjamini--Hochberg adjustment.

```{r}
study %>% 
  tidyplot(x = dose, y = score, color = group) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_pvalue(method = "wilcoxon", p.adjust.method = "BH")
```

It often makes sense to compare all experimental conditions to a control condition. For example, let's say treatment "A" is our control.

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_pvalue(ref.group = "A")
```

In some scenarios you have a mixture of significant and non-significant _p_ values.

```{r}
gene_expression %>% 
  # filter to one gene
  dplyr::filter(external_gene_name == "Apol6") %>% 
  # start plotting
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_pvalue()
```

Here you can choose to hide the non-significant _p_ value using `hide.ns = TRUE`.

```{r}
gene_expression %>% 
  # filter to one gene
  dplyr::filter(external_gene_name == "Apol6") %>% 
  # start plotting
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_pvalue(hide.ns = TRUE)
```

Finally, if you want to hide the caption with statistical information you can do this by providing `hide_info = TRUE`.

```{r}
gene_expression %>% 
  # filter to one gene
  dplyr::filter(external_gene_name == "Apol6") %>% 
  # start plotting
  tidyplot(x = condition, y = expression, color = sample_type) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_test_pvalue(hide.ns = TRUE, hide_info = TRUE)
```

There are many more things you can do with statistical comparisons. Just check out the documentation of `add_test_pvalue()` and the underlying function `ggpubr::geom_pwc()`.

# Annotation

Sometimes you wish to add annotations to provide the reader with important additional information. For example, tidyplots let's you add a `title` and a `caption`.

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_sem_errorbar() %>% 
  add_data_points() %>% 
  add_title("Interesting study") %>% 
  add_caption("Here is some more detail how the study was performed")
```

In other cases you might want to highlight specific data points or reference values in the plot. Let's take the `animals` dataset and plot `speed` versus `weight`.

```{r}
animals %>% 
  tidyplot(x = weight, y = speed) %>% 
  add_data_points()
```

Here it might be interesting to have closer at the extreme values. First, let's highlight the heaviest and the fastest animal.

```{r}
animals %>% 
  tidyplot(x = weight, y = speed) %>% 
  add_data_points() %>% 
  add_data_points(data = max_rows(weight, 1), color = "red", shape = 1, size = 2) %>% 
  add_data_points(data = max_rows(speed, 1), color = "red", shape = 1, size = 2)
```

Now it would interesting to know the names of these animals. We can plot the names of all animals.

```{r}
animals %>% 
  tidyplot(x = weight, y = speed) %>% 
  add_data_points() %>% 
  add_data_labels(label = animal)
```

Note that I provided the `label` argument to the `add_data_labels()` function to indicate the variable in the dataset that should be used for the text labels.

Another thing to note is that there is quite some overlap of labels in the lower left of the plot. Let's try to separate the data labels using the `add_data_labels_repel()` function.

```{r}
animals %>% 
  tidyplot(x = weight, y = speed) %>% 
  add_data_points() %>% 
  add_data_labels_repel(label = animal)
```

While the general idea might have been good, there are still too many labels to be plotted. So let's restrict the labels to the 3 heaviest and the 3 fastest animals.

```{r}
animals %>% 
  tidyplot(x = weight, y = speed) %>% 
  add_data_points() %>% 
  add_data_labels_repel(data = max_rows(weight, 3), animal) %>% 
  add_data_labels_repel(data = max_rows(speed, 3), animal)
```

There is lot tweaking that can be done with repelling data labels. For more details have a look at the documentation of `add_data_labels_repel()`, the underlying function `ggrepel::geom_text_repel()` and [ggrepel examples](https://ggrepel.slowkow.com/articles/examples).

As one last thing, let's add some reference lines, to highlight specific values on the x and y axis.

```{r}
animals %>% 
  tidyplot(x = weight, y = speed) %>% 
  add_reference_lines(x = 4000, y = c(100, 200)) %>% 
  add_data_points() %>% 
  add_data_labels_repel(data = max_rows(weight, 3), animal) %>% 
  add_data_labels_repel(data = max_rows(speed, 3), animal)
```

# What's more?

To dive deeper into code-based plotting, here a couple of resources.

## tidyplots documentation

- [Package index](https://jbengler.github.io/tidyplots/reference/index.html)  
Overview of all tidyplots functions

- [Get started](file:///Users/janbroderengler/GoogleDrive/R/packages/tidyplots/docs/articles/tidyplots.html)  
Getting started guide

- [Visualizing data](https://jbengler.github.io/tidyplots/articles/Visualizing-data.html)  
Article with examples for common data visualizations

- [Advanced plotting](https://jbengler.github.io/tidyplots/articles/Advanced-plotting.html)  
Article about advanced plotting techniques and workflows

- [Color schemes](https://jbengler.github.io/tidyplots/articles/Color-schemes.html)  
Article about the use of color schemes

## Other resources

- [Hands-On Programming with R](https://rstudio-education.github.io/hopr/)   
Free online book by Garrett Grolemund

- [R for Data Science](https://r4ds.hadley.nz)  
Free online book by Hadley Wickham

- [Fundamentals of Data Visualization](https://clauswilke.com/dataviz/)  
Free online book by Claus O. Wilke
