---
title: "Visualizing data"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  results = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 5,
  fig.height = 2.5,
  dpi = 300
)

```

::: {.lead}
In this article, we will explore different types of datasets and showcase the application of tidyplots to extract meaningful insights. We will cover the plotting of raw data points, amounts, and heatmaps, as well as measures for central tendency, dispersion, and uncertainty. We will conclude by visualizing distributions and proportions, and adding statistical comparisons and annotations.
:::

# Data points

Plotting the raw data points is probably the most bare bone way to visualize a dataset. The corresponding function in tidyplots is called `add_data_points()`.

```{r}
library(tidyplots)

animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points()
```

In the above example some data points appear to overlap other points. To account for this so called _overplotting_, you might want to add a thin white border around the points. This is achieved by setting the parameter `confetti = TRUE`.

```{r}
animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points(confetti = TRUE)
```

Another way is to make the points transparent using the `alpha` parameter.

```{r}
animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points(alpha = 0.3)
```

Or to change the plotting symbol to an open `shape`.

```{r}
animals %>% 
  tidyplot(x = weight, y = size) %>% 
  add_data_points(shape = 1)
```

However, data points can also be used when plotting a _discrete variable_ like `treatment` against a _continuous variable_ like `score`.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points()
```

To avoid overplotting in this scenario, there are two additional options. You can add some random noise or _jitter_ to the y position.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points_jitter()
```

Alternatively, you can use an algorithm that keeps the points centered and just moves potentially overlapping points to the sides.

```{r}
study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_data_points_beeswarm()
```

# Amounts

For some datasets, it makes sense to `count` or `sum` up data points in order to arrive to a conclusion. As one example, let's have a look at the `spendings` dataset.

```{r, results='markup'}
spendings
```

As you can see, this dataset contains family spendings over a time period of 15 days in October. Here, it might be informative to see which spending categories are reoccurring and which are just one time spendings.

```{r}
spendings %>% 
  tidyplot(x = category) %>% 
  add_count_bar()
```

One thing to note here is that the x axis labels are overlapping and are thus unreadable. There are at least two possible solutions for this. One is to swap the x and y axis.

```{r}
spendings %>% 
  tidyplot(y = category) %>% 
  add_count_bar()
```

The other one is to rotate the x axis labels.

```{r}
spendings %>% 
  tidyplot(x = category) %>% 
  add_count_bar() %>% 
  adjust_x_axis(rotate_labels = TRUE)
```

Now we can appreciate that this family had reoccurring spendings for _Food_ but just one spending for _Housing_.

Next, we ask the question how much was spend on each of the categories by plotting the `sum` amount.

```{r}
spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar() %>% 
  adjust_x_axis(rotate_labels = TRUE)
```

Note that we had to introduce the parameter `y = amount` in the `tidyplot()` function to make it clear which variable should be summed up.

I also added `color = category` in the `tidyplot()` function to have the variable `category` encoded by different colors.

Since the data labels for the variable `category` are now duplicated in the plot, one could argue that it would be justified to remove the duplicated information on the x axis.

```{r}
spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar() %>% 
  adjust_x_axis(rotate_labels = TRUE) %>% 
  remove_x_axis_labels() %>% 
  remove_x_axis_title() %>% 
  remove_x_axis_ticks()
```

Note that besides the x axis labels, I also removed the x axis ticks and x axis title to achieve a cleaner look.

Of course you are free to play around with different graphical representations of the sum values. Here is an example of a lollipop plot constructed from a thin `bar` and a `dot`.

```{r}
spendings %>% 
  tidyplot(x = category, y = amount, color = category) %>% 
  add_sum_bar(width = 0.03) %>% 
  add_sum_dot() %>% 
  add_sum_value(accuracy = 1) %>% 
  adjust_x_axis(rotate_labels = TRUE) %>% 
  remove_x_axis_labels() %>% 
  remove_x_axis_title() %>% 
  remove_x_axis_ticks()
```

I also added the sum value as text label using the `add_sum_value()` function. 

# Heatmaps

Heatmaps are a great way to plot a _continuous variable_ across to two _discrete variables_. To exemplify this, we will have a look at the `gene_expression` dataset.

```{r, results='markup'}
gene_expression %>% 
  dplyr::glimpse()
```

I used the `dplyr::glimpse()` function from the dplyr package get a nice overview of all variables including in the dataset. We will start by plotting the `expression` values of each `external_gene_name` across the `sample` variable.

```{r}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap()
```

One thing to note here is that the y axis labeks are overlapping. So let's increase the height of the plot area from 50 to 100 mm.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap() %>%
  adjust_plot_area_size(height = 100)
```

The next thing to note is that some of the rows like _Map1a_ and _Kif1a_ show very high expression while others show much lower values. Let's apply a classical technique to reserve the variations in the color for differences within each  row. This is done by calculating _row z scores_ for each row individually. Luckily, tidyplots does this for us when setting the parameter `scale = "row"` withing the `add_heatmap()` function call.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  adjust_plot_area_size(height = 100)
```

Now it much easier to appreciate the dynamics of individual genes across the samples on the x axis.

However, the rows appear to be mixed. Some having rather high expression in the "Eip" samples while others have high value in the "Hip" samples. Conveniently, in the dataset there is a variable called `direction`, which is either "up" or "down". Let's use this variable to sort our y axis.

```{r, fig.asp=0.9}
gene_expression %>% 
  tidyplot(x = sample, y = external_gene_name, color = expression) %>% 
  add_heatmap(scale = "row") %>%
  adjust_plot_area_size(height = 100) %>% 
  sort_y_axis_labels(direction)
```

# Central tendency



```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute()

energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_seaside)

energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_candy)

energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_pastel)

energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute() %>% 
  adjust_colors(colors_discrete_circle)
```

# Dispersion & uncertainty

```{r}

```

# Distribution

```{r}
distributions %>% 
  tidyplot(name, value) %>% 
  add_mean_bar(alpha = 0.3) %>% 
  add_error_bar() %>% 
  add_data_points_beeswarm()

energy %>% 
  tidyplot(x = power) %>% 
  add_histogram()

energy %>% 
  tidyplot(x = power) %>% 
  add_density_curve()

energy %>% 
  tidyplot(x = power) %>% 
  add_density_histogram() %>% 
  add_density_curve()

# multiple distributions

energy %>% 
  tidyplot(x = energy_type, y = power, color = energy_type) %>% 
  add_violin()

energy %>% 
  tidyplot(x = energy_type, y = power, color = energy_type) %>% 
  add_data_points_beeswarm(jitter_width = 0.8, alpha = 0.3)

energy %>% 
  tidyplot(x = energy_type, y = power, color = energy_type) %>% 
  add_violin() %>% 
  add_data_points_beeswarm(jitter_width = 0.8, alpha = 0.3)
```

# Proportion

```{r}
energy %>% 
  tidyplot(year, power, color = energy_source) %>% 
  add_barstack_absolute()

energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_source) %>% 
  add_pie() %>% 
  split_plot(by = year)

energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_source) %>% 
  add_donut() %>% 
  split_plot(by = year)

# pie and donut

animals %>% 
  tidyplot(y = weight, color = family) %>% 
  add_pie()

animals %>% 
  tidyplot(y = weight, color = family) %>% 
  add_donut()

animals %>% 
  tidyplot(color = family) %>% 
  add_pie()

animals %>% 
  tidyplot(color = family) %>% 
  add_donut()

# add_barstack_absolute

animals %>% 
  tidyplot(color = family) %>% 
  add_barstack_absolute()

animals %>% 
  tidyplot(x = diet, color = family) %>% 
  add_barstack_absolute()

## with y supplied

animals %>% 
  tidyplot(y = diet, color = family) %>% 
  add_barstack_absolute()

animals %>% 
  tidyplot(x = diet, color = family) %>% 
  add_barstack_absolute()

animals %>% 
  tidyplot(x = diet, y = speed, color = family) %>% 
  add_barstack_absolute()

# add_barstack_absolute

animals %>% 
  tidyplot(x = diet, color = family) %>% 
  add_barstack_relative()

animals %>% 
  tidyplot(x = diet, color = family) %>% 
  add_barstack_absolute(alpha = 0.3) %>% 
  add_areastack_absolute()

## with y supplied

animals %>% 
  tidyplot(x = diet, y = speed, color = family) %>% 
  add_barstack_absolute(alpha = 0.3) %>% 
  add_areastack_absolute()

animals %>% 
  tidyplot(y = diet, color = family) %>% 
  add_barstack_relative()

animals %>% 
  tidyplot(x = diet, y = speed, color = family) %>% 
  add_barstack_relative()
```

# Statistical comparison

```{r}

```

# Annotation

```{r}

```



```{r}
# curves

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_line() %>% 
  remove_plot_area_padding()

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_area()

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_absolute()

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_areastack_relative()

###

energy %>% 
  tidyplot(year, power, color = energy_type) %>% 
  add_barstack_absolute()

energy %>% 
  tidyplot(year, power, color = energy_type) %>% 
  add_barstack_relative()

new_colors <- c("Fossil" = "grey",
               "Nuclear" = "#F6C54D",
               "Renewable" = "#4FAE62",
               "Other" = "#C02D45")

energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_type) %>% 
  add_pie() %>% 
  adjust_colors(new_colors = new_colors) %>% 
  split_plot(by = year)

energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(y = power, color = energy_type) %>% 
  add_donut() %>% 
  adjust_colors(new_colors = new_colors) %>%  
  split_plot(by = year)

energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(x = year, y = power, color = energy_type, dodge_width = 0) %>% 
  add_sum_line() %>%
  add_sum_dot() %>% 
  adjust_colors(new_colors = new_colors)

energy %>% 
  tidyplot(x = year, y = power, color = energy_type) %>% 
  add_areastack_absolute() %>% 
  adjust_colors(new_colors = new_colors)

energy %>% 
  tidyplot(x = year, y = power, color = energy_source) %>% 
  add_areastack_absolute()

energy %>% 
  dplyr::filter(year %in% c(2005, 2010, 2015, 2020)) %>% 
  tidyplot(x = year, y = power, color = energy_type) %>% 
  add_areastack_relative() %>% 
  adjust_colors(new_colors = new_colors)

energy_week %>% 
  tidyplot(date, power, color = energy_type) %>% 
  add_sum_line() %>%
  adjust_colors(new_colors = new_colors) %>% 
  remove_plot_area_padding()

energy_week %>% 
  dplyr::summarise(power = sum(power), .by = c(date, energy_type)) %>% 
  tidyplot(date, power, color = energy_type) %>% 
  add_area() %>% 
  adjust_colors(new_colors = new_colors)

energy_week %>% 
  tidyplot(date, power, color = energy_type) %>% 
  add_areastack_absolute() %>% 
  adjust_colors(new_colors = new_colors)

energy_week %>%
  tidyplot(date, power, color = energy_type) %>% 
  add_areastack_relative() %>% 
  adjust_colors(new_colors = new_colors)

```

```{r}
study %>% 
  tidyplot(x = treatment, y = score, color = treatment) %>% 
  add_mean_dash() %>% 
  add_ci95_bar()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_ci95_ribbon() %>% 
  add_mean_line() %>% 
  add_data_points_beeswarm()

study %>% 
  tidyplot(x = treatment, y = score) %>% 
  add_ci95_ribbon() %>% 
  add_mean_line() %>% 
  add_data_points_beeswarm() %>% 
  adjust_colors("orange")

```

```{r}

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_line() %>% 
  remove_plot_area_padding()

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_area()

energy_week %>% 
  tidyplot(date, power, color = energy_source) %>% 
  add_data_points()

```

